Backward Population Synthesis: Mapping the Evolutionary History of Gravitational-Wave Progenitors

### 2022-06-04 src/tex/ms.tex
Fix GWTC-3 and hyperparameter

In full: we propose a method to ``backward" model each GW event to its
progenitor state while allowing the hyper-parameters to vary across their full
range of physical uncertainty following a two-stage process. First, we solve a
root finding problem to obtain guesses that are likely to produce the desired
system properties in the joint progenitor--hyperparameter space. We then sample
the posterior in the joint space that is induced by the DCO merger observations
using a Markov chain Monte Carlo (MCMC) algorithm that is initialized by the
roots found in the previous stage.  We demonstrate our algorithm in a realistic
setting by producing a posterior over progenitor properties and \texttt{COSMIC}
hyperparameters implied by the observation of GW150914.

We find that the progenitor properties and even ability to produce a
GW150914-like merger event is strongly correlated with hyperparameter settings;
different assumptions about the black hole masses at merger in GW150914 imply
wildly different formation channels and ZAMS progenitor masses for this event
and some nearby combinations of ZAMS masses are unable, with any reasonable
hyperparameter settings, to produce a DCO merger like GW150914.  We further
exhibit preliminary results of an ongoing analysis over the entire GWTC-3
catalog \citep{GWTC-3} that suggest that the accretion efficiency in stable mass
transfer may depend non-trivially on the primary black hole mass in merging
systems; our methodology allows for a systematic study of such dependencies,
which we will explore more fully in future work.

@@

Since the GW likelihood function
ultimately depends only on the GW parameters $\theta$, and GWTC3 already has samples over these parameters from a posterior density
\begin{equation}
    p\left( \bm{\theta} \mid d \right) \propto p\left( d \mid \bm{\theta} \right) \pi_\mathrm{GW} \left( \bm{\theta} \right),
\end{equation}
where $\pi_\mathrm{GW}$ is the prior density used for sampling, we can save the
computational cost associated with evaluating the GW likelihood by rewriting the
posterior density of $\bm{\Theta}$ in terms of the posterior density of
$\bm{\theta}$, namely

\begin{align}
    p(\bm{\Theta} | d) = \frac{p(\bm{\theta}(\bm{\Theta})| d) \pi(\bm{\Theta})}{\pi_\mathrm{GW}(\bm{\theta}(\bm{\Theta}))}.
\end{align}
We apply a kernel density estimator with a Gaussian kernel to the posterior
samples released in GWTC3 to estimate $p(\bm{\theta}(\bm{\Theta})|d)$.

@@

\noindent A small KL divergence means the reprojected \texttt{COSMIC} posterior is similar to the original posterior in the
observable space, and is thus a viable channel for that specific event. Otherwise, it either means \texttt{COSMIC}
can only explain part of the posterior or cannot explain the event at all. This is completely expected behavior since
\texttt{COSMIC}, and isolated binary evolution more generally, carries its own assumptions and is expected to
fail in reproducing a subset of the event in GWTC3. One example of this are GW events with at least one component
with mass above the pair instability supernova mass limit \citep{Woosley2017, Farmer2020}.
The reprojection could also be subject to stochasticity in the evolution of each binary
To account for this, we reproject the posterior multiple times with different random seeds and
check whether the KL divergence varies significantly. A varying KL divergence means a particular GW event is subject
to randomness in the evolution of the progenitor, and extra caution should be used when interpreting the result.

@@

\begin{figure*}[h]
    \includegraphics[width=\textwidth]{figures/GW150914_corner_zoomed_lowres.png}
    \caption{The posterior for GW150914 in both the progenitor parameter space and hyper-parameter space.
    $M_1$ and $M_2$ are the progenitors' masses. $\log{Z}$ is log metallicity at ZAMS.
    $z$ is the redshift at ZAMS.
    $\alpha$ is the common envelope efficiency.
    $f_{\rm acc}$ is the fraction of mass accreted during stable mass transfer.
    $q_{\rm crit 3}$ is the critical mass ratio on the Hertzsprung Gap,
    Note that the redshift is not fitted during the root finding process or the MCMC process.
    Once we have find the evolutionary parameters, we add the delay time to the lookback time of the observed posterior sample,
    then from the total lookback time we can compute the redshift at ZAMS.
    We highlight three panels in the corner plots to show the fine structure of the set of posterior samples in the evolutionary parameter space.
    We also color the posterior in a particular panel according to the type of parameters involved in the corner.
    Blue denotes panels that include only progenitor parameters,
    green denotes panels that include a mix of progenitor and hyperparameters,
    and orange denotes panels that include only hyperparameters.
    }
    \label{fig:GW150914_posterior}
    \script{figure1_cornerPlot.py}
\end{figure*}

Once we obtain a set of binaries which successfully map the ZAMS parameters and
hyperparameters to BBH merger masses, we re-evolve the set of ZAMS parameters
with the same physical assumptions as A21, but vary the common envelope
efficiency to explore how keeping a fixed model which only varies one
hyper-parameter contrasts to our results.
Figure~\ref{fig:compare_fixed_variable} shows the distribution of ZAMS masses
for three models (first through third columns), each with a different $\alpha$
but the same hyper-parameters as A21, as well as the results of our sampling
which allows our hyper-parameters to vary (fourth column). We find that holding
the accretion efficiency, $f_{\rm{acc}}$, and common envelope ejection
efficiency, $\alpha$, to fixed values greatly reduces the ZAMS parameter space
that produces GW150914-like mergers. In contrast, by allowing the
hyper-parameters to fully span the model uncertainty, we find that there are
distinct ZAMS parameters which produce GW150914-like mergers.

Because we can explore the full evolutionary parameter space, we can also
determine that there are ZAMS parameters which fail to produce GW150914-like
mergers \emph{regardless of our hyper-parameter choice.} For primary ZAMS masses
between $\sim95-135\,M_{\odot}$ and secondary ZAMS masses between
$\sim50-95\,M_{\odot}$, we find that there is no combination of accretion
efficiency and common envelope ejection efficiency which produces merging BBHs
that have masses consistent with GW150914's posteriors. In this region, if BBHs
with masses consistent with GW150914's masses are produced through a combination
of $\alpha$ and $f_{\rm{acc}}$, they do not merge in a Hubble time. Conversely,
in the region of the fourth column of Figure~\ref{fig:compare_fixed_variable}
which does produce GW150914-like mergers, the combination of $\alpha$,
$f_{\rm{acc}}$, and the ZAMS masses and orbital separations balance to produce
BHs with the correct masses that merge within a Hubble time.

@@

% Paragraph discussing the potential application on the population side.
To illustrate the potential benefit of using our method on the population level, we
perform the same analysis for all events in GWTC3. Plotted in figure
\ref{fig:GWTC3_f_acc_mass} is the posterior density in the $m_{1,{\rm
GW}}-f_{\rm acc}$ space for most of the events in GWTC3. Each contour represents
the $68\%$ credible interval of the posterior density for that particular event.
There is a suggestive trend showing that $f_{\rm acc}$ could increase as the mass of
the progenitor increases. Such a trend implies the stable mass transfer
phase of a less massive binary would be preferentially non-conservative, with
more conservative mass transfer in more massive binary systems.

Note that some events did not pass the KL divergence test we proposed in section
\ref{sec:method}; we do not include these systems in figure
\ref{fig:GWTC3_f_acc_mass}. Binaries that form low-mass compact objects have
lower amounts of fallback and tend to have correspondingly larger variance in
their random natal kicks. Larger kicks can unbind the progenitor binary, or lead
to wide binaries that do not merge within a Hubble time.  In these cases, the
extra variance during the reprojection can produce a posterior that may not
agree with the original posterior, hence yielding a higher KL divergence.

@@

Figure \ref{fig:GWTC3_f_acc_mass} shows that our method could in principle
reveal the correlation between progenitor parameters and hyperparameters on a
population level. Here we impose a cut to eliminate event with KL divergence larger
than 0.1.
This is a heuristic choice to exclude events that are obviously not compatible with
\texttt{COSMIC}.
A careful treatment of all events in the catalog and
discussion related to the detailed physical implications of figure
\ref{fig:GWTC3_f_acc_mass} is beyond the scope of this paper.  We defer a
detailed study of the physics related to the population of GW events to future
work.

\begin{figure}
\includegraphics[width=0.49\textwidth]{figures/GWTC3_f_acc_mass.pdf}
\caption{
    The posterior density in the $m_{1,{\rm GW}}-f_{\rm acc}$ space for most of the events in GWTC3.
    Note that $m_{1,{\rm GW}}$ here refers to the reprojected posterior instead of the posterior released by LIGO.
    Each contour is the $68\%$ credible interval of the posterior density for a particular event.
    At $m_{1,{\rm GW}} \sim 45 M_{\odot}$, the pair instability supernova mechanism prevents \texttt{COSMIC} from producing events that are more massive than this cutoff.
    Therefore, events with the majority of posterior support above this cutoff is not compatible with \texttt{COSMIC}, hence has a large KL divergence and excluded from this figure.
    On the low mass end, neutron star binaries or neutron star-black hole binaries are subjected to randomness induced by the natal kick,
    also resulting in a larger and fluctuating KL divergence, therefore they are also excluded from the analysis.
}
\label{fig:GWTC3_f_acc_mass}
\script{figure4_population.py}
\end{figure}

@@

% Single formation channel and fundamental limitation on that.
While our method allows data-driven exploration of the hyperparameters space for the first time, there are a number of improvements can be implemented in future studies.
In this study we use only \texttt{COSMIC} as our evolutionary function, which by design cannot explain all the events in GWTC3.
For example, event with either of the component mass larger than the PISN gap such as GW190521 cannot be explained by \texttt{COSMIC}.
Alternative channels such as dynamic formation channel will be needed to explain some subset of the events in GWTC3 \citep{Zevin2021}.
As GW detectors sensitivity increase, we expect to see more and more events that are unusual in some way.
Therefore, having a self-consistence population synthesis code that contains multiple formation channels is essential to accommodate the growing catalog of GW events.
In this paper, our main focus is to illustrate the concept of "back-propagating" GW events posterior sample in this work,
highlighting the capability of our method and motivating the community to build the next generation population synthesis code that can work with our method.
To avoid cluttering of focus, we discuss the physical implication of the result presented in this work under our specific assumption (i.e. using \texttt{COSMIC} as our evolutionary function) in a companion paper.

>>>>

In full: we propose a method to ``backward" model each GW event to its
progenitor state while allowing the hyperparameters to vary across their full
range of physical uncertainty following a two-stage process. First, we solve a
root finding problem to obtain guesses that are likely to produce the desired
system properties in the joint progenitor--hyperparameter space. We then sample
the posterior in the joint space that is induced by the DCO merger observations
using a Markov chain Monte Carlo (MCMC) algorithm that is initialized by the
roots found in the previous stage.  We demonstrate our algorithm in a realistic
setting by producing a posterior over progenitor properties and \texttt{COSMIC}
hyperparameters implied by the observation of GW150914.

We find that the progenitor properties and even the ability to produce a
GW150914-like merger event is strongly correlated with hyperparameter settings;
different assumptions about the black hole masses at merger in GW150914 imply
wildly different formation channels and ZAMS progenitor masses for this event
and some nearby combinations of ZAMS masses are unable, with any reasonable
hyperparameter settings, to produce a DCO merger like GW150914.  We further
exhibit preliminary results of an ongoing analysis over the entire GWTC-3
catalog \citep{GWTC-3} that suggest that the accretion efficiency in stable mass
transfer may depend non-trivially on the primary black hole mass in merging
systems; our methodology allows for a systematic study of such dependencies,
which we will explore more fully in future work.

@@

Since the GW likelihood function
ultimately depends only on the GW parameters $\theta$, and GWTC-3 already has samples over these parameters from a posterior density
\begin{equation}
    p\left( \bm{\theta} \mid d \right) \propto p\left( d \mid \bm{\theta} \right) \pi_\mathrm{GW} \left( \bm{\theta} \right),
\end{equation}
where $\pi_\mathrm{GW}$ is the prior density used for sampling, we can save the
computational cost associated with evaluating the GW likelihood by rewriting the
posterior density of $\bm{\Theta}$ in terms of the posterior density of
$\bm{\theta}$, namely

\begin{align}
    p(\bm{\Theta} | d) = \frac{p(\bm{\theta}(\bm{\Theta})| d) \pi(\bm{\Theta})}{\pi_\mathrm{GW}(\bm{\theta}(\bm{\Theta}))}.
\end{align}
We apply a kernel density estimator with a Gaussian kernel to the posterior
samples released in GWTC-3 to estimate $p(\bm{\theta}(\bm{\Theta})|d)$.

@@

\noindent A small KL divergence means the reprojected \texttt{COSMIC} posterior is similar to the original posterior in the
observable space, and is thus a viable channel for that specific event. Otherwise, it either means \texttt{COSMIC}
can only explain part of the posterior or cannot explain the event at all. This is completely expected behavior since
\texttt{COSMIC}, and isolated binary evolution more generally, carries its own assumptions and is expected to
fail in reproducing a subset of the event in GWTC-3. One example of this are GW events with at least one component
with mass above the pair instability supernova mass limit \citep{Woosley2017, Farmer2020}.
The reprojection could also be subject to stochasticity in the evolution of each binary
To account for this, we reproject the posterior multiple times with different random seeds and
check whether the KL divergence varies significantly. A varying KL divergence means a particular GW event is subject
to randomness in the evolution of the progenitor, and extra caution should be used when interpreting the result.

@@

\begin{figure*}[h]
    \includegraphics[width=\textwidth]{figures/GW150914_corner_zoomed_lowres.png}
    \caption{The posterior for GW150914 in both the progenitor parameter space and hyperparameter space.
    $M_1$ and $M_2$ are the progenitors' masses. $\log{Z}$ is log metallicity at ZAMS.
    $z$ is the redshift at ZAMS.
    $\alpha$ is the common envelope efficiency.
    $f_{\rm acc}$ is the fraction of mass accreted during stable mass transfer.
    $q_{\rm crit 3}$ is the critical mass ratio on the Hertzsprung Gap,
    Note that the redshift is not fitted during the root finding process or the MCMC process.
    Once we have find the evolutionary parameters, we add the delay time to the lookback time of the observed posterior sample,
    then from the total lookback time we can compute the redshift at ZAMS.
    We highlight three panels in the corner plots to show the fine structure of the set of posterior samples in the evolutionary parameter space.
    We also color the posterior in a particular panel according to the type of parameters involved in the corner.
    Blue denotes panels that include only progenitor parameters,
    green denotes panels that include a mix of progenitor and hyperparameters,
    and orange denotes panels that include only hyperparameters.
    }
    \label{fig:GW150914_posterior}
    \script{figure1_cornerPlot.py}
\end{figure*}

Once we obtain a set of binaries which successfully map the ZAMS parameters and
hyperparameters to BBH merger masses, we re-evolve the set of ZAMS parameters
with the same physical assumptions as A21, but vary the common envelope
efficiency to explore how keeping a fixed model which only varies one
hyperparameter contrasts to our results.
Figure~\ref{fig:compare_fixed_variable} shows the distribution of ZAMS masses
for three models (first through third columns), each with a different $\alpha$
but the same hyperparameters as A21, as well as the results of our sampling
which allows our hyperparameters to vary (fourth column). We find that holding
the accretion efficiency, $f_{\rm{acc}}$, and common envelope ejection
efficiency, $\alpha$, to fixed values greatly reduces the ZAMS parameter space
that produces GW150914-like mergers. In contrast, by allowing the
hyperparameters to fully span the model uncertainty, we find that there are
distinct ZAMS parameters which produce GW150914-like mergers.

Because we can explore the full evolutionary parameter space, we can also
determine that there are ZAMS parameters which fail to produce GW150914-like
mergers \emph{regardless of our hyperparameter choice.} For primary ZAMS masses
between $\sim95-135\,M_{\odot}$ and secondary ZAMS masses between
$\sim50-95\,M_{\odot}$, we find that there is no combination of accretion
efficiency and common envelope ejection efficiency which produces merging BBHs
that have masses consistent with GW150914's posteriors. In this region, if BBHs
with masses consistent with GW150914's masses are produced through a combination
of $\alpha$ and $f_{\rm{acc}}$, they do not merge in a Hubble time. Conversely,
in the region of the fourth column of Figure~\ref{fig:compare_fixed_variable}
which does produce GW150914-like mergers, the combination of $\alpha$,
$f_{\rm{acc}}$, and the ZAMS masses and orbital separations balance to produce
BHs with the correct masses that merge within a Hubble time.

@@

% Paragraph discussing the potential application on the population side.
To illustrate the potential benefit of using our method on the population level, we
perform the same analysis for all events in GWTC-3. Plotted in figure
\ref{fig:GWTC-3_f_acc_mass} is the posterior density in the $m_{1,{\rm
GW}}-f_{\rm acc}$ space for most of the events in GWTC-3. Each contour represents
the $68\%$ credible interval of the posterior density for that particular event.
There is a suggestive trend showing that $f_{\rm acc}$ could increase as the mass of
the progenitor increases. Such a trend implies the stable mass transfer
phase of a less massive binary would be preferentially non-conservative, with
more conservative mass transfer in more massive binary systems.

Note that some events did not pass the KL divergence test we proposed in section
\ref{sec:method}; we do not include these systems in figure
\ref{fig:GWTC-3_f_acc_mass}. Binaries that form low-mass compact objects have
lower amounts of fallback and tend to have correspondingly larger variance in
their random natal kicks. Larger kicks can unbind the progenitor binary, or lead
to wide binaries that do not merge within a Hubble time.  In these cases, the
extra variance during the reprojection can produce a posterior that may not
agree with the original posterior, hence yielding a higher KL divergence.

@@

Figure \ref{fig:GWTC-3_f_acc_mass} shows that our method could in principle
reveal the correlation between progenitor parameters and hyperparameters on a
population level. Here we impose a cut to eliminate event with KL divergence larger
than 0.1.
This is a heuristic choice to exclude events that are obviously not compatible with
\texttt{COSMIC}.
A careful treatment of all events in the catalog and
discussion related to the detailed physical implications of figure
\ref{fig:GWTC-3_f_acc_mass} is beyond the scope of this paper.  We defer a
detailed study of the physics related to the population of GW events to future
work.

\begin{figure}
\includegraphics[width=0.49\textwidth]{figures/GWTC-3_f_acc_mass.pdf}
\caption{
    The posterior density in the $m_{1,{\rm GW}}-f_{\rm acc}$ space for most of the events in GWTC-3.
    Note that $m_{1,{\rm GW}}$ here refers to the reprojected posterior instead of the posterior released by LIGO.
    Each contour is the $68\%$ credible interval of the posterior density for a particular event.
    At $m_{1,{\rm GW}} \sim 45 M_{\odot}$, the pair instability supernova mechanism prevents \texttt{COSMIC} from producing events that are more massive than this cutoff.
    Therefore, events with the majority of posterior support above this cutoff is not compatible with \texttt{COSMIC}, hence has a large KL divergence and excluded from this figure.
    On the low mass end, neutron star binaries or neutron star-black hole binaries are subjected to randomness induced by the natal kick,
    also resulting in a larger and fluctuating KL divergence, therefore they are also excluded from the analysis.
}
\label{fig:GWTC-3_f_acc_mass}
\script{figure4_population.py}
\end{figure}

@@

% Single formation channel and fundamental limitation on that.
While our method allows data-driven exploration of the hyperparameters space for the first time, there are a number of improvements can be implemented in future studies.
In this study we use only \texttt{COSMIC} as our evolutionary function, which by design cannot explain all the events in GWTC-3.
For example, event with either of the component mass larger than the PISN gap such as GW190521 cannot be explained by \texttt{COSMIC}.
Alternative channels such as dynamic formation channel will be needed to explain some subset of the events in GWTC-3 \citep{Zevin2021}.
As GW detectors sensitivity increase, we expect to see more and more events that are unusual in some way.
Therefore, having a self-consistence population synthesis code that contains multiple formation channels is essential to accommodate the growing catalog of GW events.
In this paper, our main focus is to illustrate the concept of "back-propagating" GW events posterior sample in this work,
highlighting the capability of our method and motivating the community to build the next generation population synthesis code that can work with our method.
To avoid cluttering of focus, we discuss the physical implication of the result presented in this work under our specific assumption (i.e. using \texttt{COSMIC} as our evolutionary function) in a companion paper.
